\bf Excuse me, do you have a moment to talk about version control?

### 2017-05-19 README.tex
Revisions, citations, links

Git has been re-purposed by the data science community. In addition to
using it for source code, we use it to manage the motley collection of
files that make up typical data analytical projects, which often consist
of data, figures, reports, and, yes, source code. Even those who
identify more as statistician than data scientist generally have a
similar mix of files that are the artifacts of a project.

@@

The Git way is to track the evolution of that file, through a series of
commits, each equipped with an explanatory message and a nickname. All
collaborators sync regularly to a common version, acknowledging that the
difficulty of merging goes up faster than the size of the difference
(cite good enough). Especially important versions get a human-readable
tag, to signal a meaningful milestone. Yes, there is some pain in
adopting the formalism of Git, but it is worth it.

\emph{another good image to emulate, from the Bartlett Git for Designers
talk} \includegraphics{bartlett-copies-vs-commits.png}

@@

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  When you are new with Git (or, frankly, even when you're not), it's
  common to goof up the Git infrastructure for a project. Note that your
  files can absolutely be intact and safe, even while the Git tracking
  is a bit confused. Of course there are official Git remedies, but
  sometimes the easiest fix is to clone a fresh copy from GitHub, patch
  things up with the changes that only exist locally, and move on with
  your life. This workaround obviously requires the existence of a
  recent copy on GitHub.
\item
  The highly functional web interfaces mentioned above are often the
  most pleasant and natural way to navigate and search your files, even
  though all the same information exists locally. It is a pleasure to
  browse through your own work, across multiple projects or files and
  across time, as if it's a well-designed website. You must push your
  work to GitHub to enjoy this.
\end{enumerate}

GitHub \textbf{issues} are another powerful feature of the platform.
Recall that we are repurposing Git, a tool designed to facilitate
software development. The issues for a project are its bug tracker. For
projects that are not pure software development, we co-opt this
machinery to organize our to do list more generally. The basic unit is
an issue and you can interact with them in two ways.

@@

This is one-time or once-per-computer setup.

\begin{itemize}
\item
  Register for a free account with GitHub.
\item
  Install Git. Depending on your OS, Git might already be installed. But
  many of us will need to install it or might choose to update to the
  most recent version. Some basic configuration is critical, such as
  setting your username and email.
\item
  Install a local Git client. Optional but highly recommended. A Git
  client is software that provides a graphical user interface for Git,
  which is otherwise command-line only. If you are an R user, you will
  find that RStudio provides a great deal of this functionality. There
  are some notable gaps, however, so you might still choose to install a
  dedicated and comprehensive Git client such as SourceTree or
  GitKraken. Git is a file-based system, so you can do some operations
  from RStudio, others from SourceTree, and others from the shell.
\item
  Confirm, with a practice repository, that local Git can send and
  receive the current version of the repository on GitHub, known as
  \emph{pushing} and \emph{pulling}, respectively. This will require
  authenticating yourself with GitHub from a local shell or Git client.
  At this point, most people elect to do a bit of extra setup to ensure
  that they are not repeatedly challenged for their GitHub credentials
  going forward.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  I repeat: consider using a graphical front-end for Git, a.k.a. a ``Git
  client'', versus restricting yourself to the command line interface.
\item
  Establish confidence in the basics (e.g.~make a change, commit it,
  push it) before wading into more advanced usage (e.g.~branching).
\item
  Commit yourself to Git usage on a project that will provide sustained
  practice over several months. Usage in a course is great, because it
  provides a relentless stream of small deadlines.
\item
  Realize that no one is giving out Git style points. It's ok to
  ``power-cycle'', i.e.~re-initialize the Git repository, to get
  unstuck.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Dedicate a local directory (a.k.a ``folder'') to it.
\item
  Make it an RStudio Project. Optional but recommended; obviously only
  applies to projects involving R and users of RStudio.
\item
  Make it a Git repository.
\end{itemize}

This is once-per-project setup and can happen at project inception or at
any later point. Chances are your project already lives in a dedicated
directory. Making this directory an RStudio Project and Git repository
boils down to allowing those applications to leave notes for themselves
in hidden files or directories. The project is still a regular directory
on your computer, that you can locate, name, move, and generally
interact with as you wish. You don't have to handle it with special
gloves!

@@

\begin{itemize}
\tightlist
\item
  Go about your usual business, e.g.~writing R scripts or authoring
  reports in LaTeX or R Markdown. But instead of only \emph{saving}
  individual files, periodically you make a \textbf{commit}, which takes
  a snapshot of all the files in the entire project.

  \begin{itemize}
  \tightlist
  \item
    Have you ever versioned a file
    \href{http://www.phdcomics.com/comics/archive.php?comicid=1531}{by
    adding your initials or the date}? That is effectively a
    \textbf{commit}, albeit only for a single file: it is a version that
    is significant to you and that you might want to inspect or revert
    to later.
  \end{itemize}
\item
  Push commits to GitHub periodically.

@@

This is a moderate change to your normal, daily workflow. It feels weird
at first, but quickly becomes second nature. In
\href{http://stat545.com}{STAT 545} students are required to submit all
coursework via GitHub, starting in week one. Most have never seen Git
before and do not identify as ``programmers''. It is a major topic in
class and office hours for the first two weeks. Then we practically
never discuss it again.

\emph{I am not saying anything about pulling and collaboration here,
which seems a bit extreme, even if I can't say much}

@@

Recall that a repository -- or ``repo'' -- is just a directory of files
that Git is going to manage holistically. A commit functions like a
snapshot of all the files in the repo, at a specific moment. Under the
hood, that is not exactly how Git implements things. Mental models don't
have to be accurate in order to be useful, but in this case there's some
value in aligning the two.

@@

\begin{figure}
\centering
\includegraphics{bartlett-commit-history.png}
\caption{}
\end{figure}

@@

Every time you make a commit you must also write a short \emph{commit
message}. Ideally, this conveys the \emph{motivation} for the change.
Remember, the diff will show the content. When you revisit a project
after a break or need to digest recent changes made by a colleague,
looking at the \emph{history}, by reading commit messages and skimming
through diffs, is an extremely efficient way to get up to speed.

Every commit needs some sort of nickname, so you can identify it. Git
does this automatically, assigning each commit what is called a
\emph{SHA}, a seemingly random string of 40 letters and numbers (it is
not, in fact, random but is a SHA-1 checksum hash of the commit). Though
you will be exposed to these, you don't have to handle them directly
very often and, when you do, usually the first 7 characters suffice. You
can also designate certain snapshots as special with a \emph{tag}, which
is a name of your choosing. In a software project, it is typical to tag
a release with its version, e.g., ``v1.0.3''. For a manuscript or
analytical project, you might tag the version submitted to a journal or
transmitted to external collaborators.

@@

This may seem unrelated to Git, GitHub, and R, but we need to talk about
Markdown. Markdown is a markup language, like HTML and LaTeX, but
designed to be as lightweight as possible. The goal is still to separate
form and content, but also to prioritize human-readability, even at the
cost of fancy features. Markdown is in wide use on sites like Wikipedia,
WordPress, StackOverflow, and, yes, GitHub. These sites use Markdown
because it allows a diverse population of site users to author
decent-looking web content, with hyperlinks and some formatting. Do not
build this up into some heroic, LaTeX-level learning task, for it is
not. If you can write an email, you can write Markdown.

@@

Markdown is extra special for R users because of R Markdown, which is
just Markdown that includes chunks of R code. Again, do not regard R
Markdown as something you must clear your schedule to learn. If you can
write email and a bit of R code, you can write R Markdown. The rmarkdown
package converts R Markdown (\texttt{.Rmd} files) to Markdown
(\texttt{.md} files), running the code and inserting the results,
including figures, into the document. This is powered by another
package, knitr, under the hood. This process is made especially easy in
RStudio, but is by no means limited to users of that application.

@@

You do not, in fact, even need to work in R Markdown to exploit this. It
works with plain R scripts as well. You can use exactly the same
machinery to prepare a rendered version of an R script, i.e.~to go from
\texttt{.R} to \texttt{.md}. Again, RStudio makes this especially easy,
but this is not limited to RStudio. Once the Markdown file is pushed to
GitHub, it is as if the reader has run your code or is able to look over
your shoulder at your R session. This provides an lightweight system for
exposing work-in-progress to collaborators, without slowing down to
create separate reports. Comment lines that begin with
\texttt{\#\textquotesingle{}} are elevated to top-level prose, providing
a way to make the document more welcoming for a reader. Once there are
many prose comments, you might decide to switch from \texttt{.R} to
\texttt{.Rmd}, have proper top-level prose, and move the code down into
chunks.

@@

Let's zoom back out again and consider R Markdown and the rmarkdown
package more generally. I want to point out that R Markdown can be
rendered to many more formats than Markdown! I have emphasized the
production of Markdown here because it is extremely useful in
GitHub-hosted projects and that seems to be underappreciated. But the
rmarkdown package can convert \texttt{.Rmd} to a wide array of output
formats, including HTML, PDF, and Word (\texttt{.docx}). The Markdown
must be created regardless. Sometimes it is all you need, such as on
GitHub. In other contexts, it is just a necessary intermediate and may
even be discarded.

@@

\begin{itemize}
\tightlist
\item
  R markdown \texttt{.Rmd} --\textgreater{} markdown \texttt{.md}
\item
  R markdown \texttt{.Rmd} --\textgreater{} markdown \texttt{.md}
  --\textgreater{} \texttt{.html} or \texttt{.pdf} or \texttt{.docx}
\item
  R script \texttt{.R} --\textgreater{} results as \texttt{.csv} or
  \texttt{.rds} and figures \texttt{.png}
\item
  LaTeX \texttt{.tex} --\textgreater{} WHAT GOES HERE NOW?
  --\textgreater{} \texttt{.pdf}
\end{itemize}

The files at the far left are clearly source files. In the case of an R
script, this is literally true, but it's morally true for R markdown and
LaTeX files too. These are files that you directly create and edit ``by
hand''.

The files at the far right are clearly derived and are often described
as ``targets''. These files are programmatically generated from source
files (and possibly other inputs). These files are the product and they
have external value, often for communicating ideas and results.

@@

A critical issue for workflow happiness is figuring out how to manage
the production and storage of source, intermediates, and targets with
respect to Git. You can direct Git to ignore specific files or certain
types of files, such as autosaves created by your editor. This reduces
clutter in your project: Git will not pester you to commit changes to
these files and they will not appear in the associated GitHub
repository. A file that Git does not ignore is said to be
\emph{tracked}.

@@

@@

\emph{Will it make collaboration harder?} \emph{talk about the scenario
Nick brought up re: binary files like PDF being a common source of merge
conflict
\url{https://github.com/dsscollection/git-github-for-stats/issues/6}}

@@

Managing a project via Git/GitHub is much more like the Google Doc
scenario, but also offers some of the attractive features of ``edit,
save, attach''. With Git/GitHub, collaborators can work offline and
there can be independent lines of development. The real power comes from
regular and structured reconciliation of all the files in the project
via Git/GitHub. It is definitely more complicated than collaborating on
a Google Doc, but also more powerful.

@@

\begin{figure}
\centering
\includegraphics{bartlett-merge-commit.png}
\caption{}
\end{figure}

\begin{figure}
\centering
\includegraphics{bartlett-pull.png}
\caption{}
\end{figure}

@@

Transparency about \citep{greenwade93} process and product is
increasingly important in science. The SOMETHING for reproducibility is
well accepted. A more underappreciated benefit is democratization of our
field, as this affords a much broader audience a clear view of how
scientists and programmers work.

@@

\href{https://www.rstudio.com/products/rstudio-desktop/}{RStudio IDE}.

>>>>

Git has been re-purposed by the data science community
\citep{Ram2013, git-for-humans}. In addition to using it for source
code, we use it to manage the motley collection of files that make up
typical data analytical projects, which often consist of data, figures,
reports, and, yes, source code. Even those who identify more as
statistician than data scientist generally have a similar mix of files
that are the artifacts of a project.

@@

The Git way is to track the evolution of that file, through a series of
commits, each equipped with an explanatory message and a nickname. All
collaborators sync regularly to a common version, acknowledging that the
difficulty of merging goes up faster than the size of the difference
\citep{good-enough}. Especially important versions get a human-readable
tag, to signal a meaningful milestone. Yes, there is some pain in
adopting the formalism of Git, but it is worth it.

\emph{make a figure related to the above (reference those file names?)
but juxtaposing different workflows, inspiration from the Bartlett Git
for Designers talk}

\includegraphics[width=1\linewidth]{bartlett-copies-vs-commits}

@@

\emph{possible images include a modified version of this atlassian image
(no arrows between people, yes arrow from people to github), somehow
connected to a screenshot of a GitHub repo}

\includegraphics[width=1\linewidth]{atlassian-git-tutorials-syncing}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  When you are new with Git (or, frankly, even when you're not), it's
  common to goof up the Git infrastructure for a project. Note that your
  files can be intact and safe, even while the Git tracking is a bit
  confused. Of course there are official Git remedies, but sometimes the
  easiest fix is to clone a fresh copy from GitHub, patch things up with
  the changes that only exist locally, and move on with your life. This
  workaround obviously requires the existence of a recent copy on
  GitHub.
\item
  The highly functional web interfaces mentioned above are often the
  most pleasant and natural way to navigate and search your files, even
  though all the same information exists locally. It is a pleasure to
  browse through your own work, across multiple projects or files and
  across time, as if it's a well-designed website. You must push your
  work to GitHub to enjoy this.
\end{enumerate}

\textbf{GitHub issues} are another powerful feature of the platform.
Recall that we are repurposing Git, a tool designed to facilitate
software development. The issues for a project are its bug tracker. For
projects that are not pure software development, we co-opt this
machinery to organize our to do list more generally. The basic unit is
an issue and you can interact with one in two ways.

@@

Here is the initial setup process, which you do once or, for some steps,
once per computer.

\begin{itemize}
\item
  Register for a free account with GitHub.
\item
  Install Git. Depending on your OS, Git might already be installed. But
  many of us may need to install it or will choose to update to a more
  recent version. Some basic configuration is critical, such as setting
  your username and email.
\item
  Install a local Git client \emph{(optional but highly recommended}). A
  Git client is software that provides a graphical user interface for
  Git, which is otherwise command-line only. If you are an R user, you
  will find that
  \href{https://www.rstudio.com/products/rstudio-desktop/}{RStudio}
  provides a great deal of this functionality. There are some notable
  gaps, however, so you might still choose to install a dedicated and
  comprehensive Git client such as
  \href{https://www.sourcetreeapp.com}{SourceTree} or
  \href{https://www.gitkraken.com}{GitKraken}. Git is a file-based
  system, so you can do some operations from RStudio, others from
  SourceTree, and others from the shell.
\item
  Confirm, with a practice repository, that local Git can send and
  receive the current version of the repository on GitHub, known as
  \textbf{pushing} and \textbf{pulling}, respectively. This will require
  authenticating yourself with GitHub from a local shell or Git client.
  At this point, most people elect to do a bit of extra setup to ensure
  that they are not repeatedly challenged for their GitHub credentials
  going forward.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  I repeat: consider using a graphical front-end for Git, a.k.a. a Git
  client, versus restricting yourself to the command line interface.
\item
  Establish confidence in the basics (e.g.~make a change, commit it,
  push it) before wading into more advanced usage (e.g.~branching).
\item
  Commit yourself to Git usage on a project that will provide sustained
  practice over several months. Usage in a course is great, because it
  provides a relentless stream of small deadlines.
\item
  Realize that no one is giving out Git style points. It's ok to
  ``power-cycle'', i.e.~re-initialize the Git repository, to get
  unstuck.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Dedicate a local directory or folder to it.
\item
  Make it an RStudio Project. \emph{Optional but recommended; obviously
  only applies to projects involving R and users of RStudio.}
\item
  Make it a Git repository.
\end{itemize}

This setup happens once per project and can happen at project inception
or at any later point. Chances are your project already lives in a
dedicated directory. Making this directory an RStudio Project and Git
repository boils down to allowing those applications to leave notes for
themselves in hidden files or directories. The project is still a
regular directory on your computer, that you can locate, name, move, and
generally interact with as you wish. You don't have to handle it with
special gloves!

@@

\begin{itemize}
\tightlist
\item
  Go about your usual business, writing R scripts or authoring reports
  in LaTeX or R Markdown. But instead of only \emph{saving} individual
  files, periodically you make a \textbf{commit}, which takes a snapshot
  of all the files in the entire project.

  \begin{itemize}
  \tightlist
  \item
    If you have ever versioned a file
    \href{http://www.phdcomics.com/comics/archive.php?comicid=1531}{by
    adding your initials or the date}, you have effectively made a
    commit, albeit only for a single file. It is a version that is
    significant to you and that you might want to inspect or revert to
    later.
  \end{itemize}
\item
  Push commits to GitHub periodically.

@@

This is a moderate change to your normal, daily workflow. It feels weird
at first, but quickly becomes second nature. In
\href{http://stat545.com}{STAT 545} students are required to submit all
coursework via GitHub, starting in week one. Most have never seen Git
before and do not identify as programmers. It is a major topic in class
and office hours for the first two weeks. Then we practically never
discuss it again.

@@

Recall that a repository or repo is just a directory of files that Git
will manage holistically. A commit functions like a snapshot of all the
files in the repo, at a specific moment. Under the hood, that is not
exactly how Git implements things. Mental models don't have to be
accurate in order to be useful, but in this case there's some value in
aligning the two.

@@

\includegraphics[width=1\linewidth]{bartlett-commit-history}

@@

Every time you make a commit you must also write a short \textbf{commit
message}. Ideally, this conveys the motivation for the change. Remember,
the diff will show the content. When you revisit a project after a break
or need to digest recent changes made by a colleague, looking at the
\textbf{history}, by reading commit messages and skimming through diffs,
is an extremely efficient way to get up to speed.

Every commit needs some sort of nickname, so you can identify it. Git
does this automatically, assigning each commit what is called a SHA, a
seemingly random string of 40 letters and numbers (it is not, in fact,
random but is a SHA-1 checksum hash of the commit). Though you will be
exposed to these, you don't have to handle them directly very often and,
when you do, usually the first 7 characters suffice. You can also
designate certain snapshots as special with a \textbf{tag}, which is a
name of your choosing. In a software project, it is typical to tag a
release with its version, e.g., ``v1.0.3''. For a manuscript or
analytical project, you might tag the version submitted to a journal or
transmitted to external collaborators.

@@

This may seem unrelated to Git, GitHub, and R, but we need to talk about
\href{https://daringfireball.net/projects/markdown/syntax}{Markdown}.
Markdown is a markup language, like HTML and LaTeX, but designed to be
as lightweight as possible. The goal is still to separate form and
content, but also to prioritize human-readability, even at the cost of
fancy features. Markdown is in wide use on sites like
\href{https://en.support.wordpress.com/markdown/}{WordPress},
\href{https://stackoverflow.com/editing-help}{StackOverflow}, and, yes,
\href{https://help.github.com/categories/writing-on-github/}{GitHub}.
These sites use Markdown because it allows a diverse population of site
users to author decent-looking web content, with hyperlinks and some
formatting. Do not build this up into some heroic, LaTeX-level learning
task, for it is not. If you can write an email, you can write Markdown.

@@

Markdown is extra special for R users because of
\href{http://rmarkdown.rstudio.com}{R Markdown}, which is just Markdown
that includes chunks of R code. Again, do not regard R Markdown as
something you must clear your schedule to learn. If you can write email
and a bit of R code, you can write R Markdown. The
\href{https://CRAN.R-project.org/package=rmarkdown}{rmarkdown package}
\citep{rmd-pkg} converts R Markdown (\texttt{.Rmd} files) to Markdown
(\texttt{.md} files), running the code and inserting the results,
including figures, into the document. This is powered by another
package, \href{https://CRAN.R-project.org/package=knitr}{knitr}
\citep{knitr-pkg, knitr-book}, under the hood. This process is made
especially easy in RStudio, but is by no means limited to users of that
application.

@@

You do not, in fact, even need to work in R Markdown to exploit this. It
\href{http://rmarkdown.rstudio.com/articles_report_from_r_script.html}{works
with plain R scripts} as well. You can use exactly the same machinery to
prepare a rendered version of an R script, i.e.~to go from \texttt{.R}
to \texttt{.md}. Again, RStudio makes this especially easy, but this is
not limited to RStudio. Once the Markdown file is pushed to GitHub, it
is as if the reader has run your code or is able to look over your
shoulder at your R session. This provides an lightweight system for
exposing work-in-progress to collaborators, without slowing down to
create separate reports. Comment lines that begin with
\texttt{\#\textquotesingle{}} are elevated to top-level prose, providing
a way to make the document more welcoming for a reader. Once there are
many prose comments, you might decide to switch from \texttt{.R} to
\texttt{.Rmd}, have proper top-level prose, and move the code down into
chunks.

@@

Let's zoom back out again and consider R Markdown and the rmarkdown
package more generally. I want to point out that R Markdown can be
rendered to many more formats than Markdown! I have emphasized the
production of Markdown here because it is extremely useful in
GitHub-hosted projects and that seems to be underappreciated. But the
rmarkdown package can convert \texttt{.Rmd} to a wide array of
\href{http://rmarkdown.rstudio.com/lesson-9.html}{output formats},
including HTML, PDF, and Word (\texttt{.docx}). The Markdown must be
created regardless. Sometimes it is all you need, such as on GitHub. In
other contexts, it is just a necessary intermediate and may even be
discarded.

@@

\begin{itemize}
\tightlist
\item
  R markdown \texttt{.Rmd} --\textgreater{} markdown \texttt{.md}
\item
  R markdown \texttt{.Rmd} --\textgreater{} markdown \texttt{.md}
  --\textgreater{} \texttt{.html} or \texttt{.pdf} or \texttt{.docx}
\item
  R script \texttt{.R} --\textgreater{} results as \texttt{.csv} or
  \texttt{.rds} and figures \texttt{.png}
\item
  LaTeX \texttt{.tex} --\textgreater{} \texttt{.aux}, \texttt{.bbl},
  etc. --\textgreater{} \texttt{.pdf}
\end{itemize}

The files at the far left are clearly source files. In the case of an R
script, this is literally true, but it's morally true for R markdown and
LaTeX files too. These are files that you directly create and edit by
hand.

The files at the far right are clearly derived and are often described
as \textbf{targets}. These files are programmatically generated from
source files (and possibly other inputs). These files are the product
and they have external value, often for communicating ideas and results.

@@

A critical issue for workflow happiness is figuring out how to manage
the production and storage of source, intermediates, and targets with
respect to Git. You can direct Git to ignore specific files or certain
types of files, such as autosaves created by your editor. This reduces
clutter in your project: Git will not pester you to commit changes to
these files and they will not appear in the associated GitHub
repository. A file that Git does not ignore is said to be
\textbf{tracked}.

@@

\emph{I suspect this is a good place to be less wordy? I am so
frustrated with people doing stupid things here, though!}

@@

\emph{Will it make collaboration harder?} Prose to be written: Talk
about the scenario Nick brought up re: binary files like PDF being a
common source of merge conflict
\url{https://github.com/dsscollection/git-github-for-stats/issues/6}. I
have a solution! Stop using PDF as your default output format! Make
Markdown the output format during development. Your Git/GitHub problems
just went away. You're welcome.

@@

Managing a project via Git/GitHub is much more like the Google Doc
scenario, but also offers some of the attractive features of ``edit,
save, attach''. With Git/GitHub, collaborators can work offline and
there can be independent lines of development. The real power comes from
regular and structured reconciliation of all versions of all the files
in the project via Git/GitHub. It is definitely more complicated than
collaborating on a Google Doc, but also more powerful.

@@

\includegraphics[width=1\linewidth]{bartlett-merge-commit}

\includegraphics[width=1\linewidth]{bartlett-pull}

@@

Transparency about process and product is increasingly important in
science. The SOMETHING for reproducibility is well accepted. A more
underappreciated benefit is democratization of our field, as this
affords a much broader audience a clear view of how scientists and
programmers work.

@@
